Visual Analytics of Large Bipartite Networks assisted by Multilevel Strategies\thanks{Supported by FAPESP, project 2017/05838-3}

### 2019-05-22 paper.tex
retoques de escrita

\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information for the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities~\cite{alan1}.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or may be considered as exhibiting bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
Their employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks have been reported, but their exploitation for visualizing bipartite networks
was not found by the authors.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presented a simplified version of the network for the user, which then
requests for supervertices (or collections of them) to be uncoarsened and presented in more detail.

This paper is organized as follows: in Section~\ref{rel} the related work is examined, while in~\ref{nom} are selected remarks about the vocabulary.
The method is delineated in Section~\ref{des} and
the software implementation is then described in Section~\ref{sof}.
Results and discussion are in~\ref{res}.
Finally, Section~\ref{con} holds concluding and further work statements.

\subsection{Related work}\label{rel}
%%%
% artigos que devemos citar pelo histórico científico
% da estratégia ML, da visualização ML
% e os que devemos citar pelo ML em redes bipartidas
% e desenvolvidos no ICMC
Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
Also, the aggregation of clusters have been reported, and comprises an approach that resembles
the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
Even so, the authors are not aware of previous reports on the use of multilevel strategies
for the visualization and navigation of bipartite networks.
Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system for visual analytics, i.e. the other works do not
address data analysis by interactive visual interfaces,
including all multilevel strategies for visualization.

\subsection{Nomenclature remarks}\label{nom}
%%%
% level vs layer
% simple and heterogeneous graphs
% multiplex, ?
% multilevel strategies for optimization
% supernode metanode, 
The main vocabulary issue that arises in the context of this article is between
level and layer. Layers are the sets of a bipartite network, while levels are the sequence
of coarsened networks.
Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there
are more then one type of node.
Bipartite networks may be regarded as an elementary case of heterogeneous networks,
but it has one additional restriction: only nodes of different types are connected.
Another usual concept in this context is that of a multilayer network, in which there are layers,
i.e. partitions of nodes, in addition to nodes and links.
Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layers.
Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
A thorough exposition of the vocabulary are beyond the scope of this article, but
some attention for the issue is helpful to assist searches and newcomers.

@@

A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network.
The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
from the matched nodes.
There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
developed for bipartite networks~\cite{}.
A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks, but because it has yielded simplified networks which carries essential topological features of the original networks, and that result is perceived is the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.

>>>>

\keywords{Network visualization \and Multilevel strategies \and Visual analytics \and Big data \and Complex networks \and Data visualization.}
\end{abstract}
%
\section{Introduction}
The visualization of large-scale networks poses challenges both in terms of computational costs
and of effective presentation of the information to the user~\cite{tang,staudt}.
These issues may be aggravated in the case of bipartite networks,
due to their sparsity and topological complexities~\cite{alan1}.
Bipartite networks are comprised of two partitions of nodes,
called ``layers'',
and links are not incident between nodes in the same partition.
Such network type arises very often and naturally from the representation
of relations among two kinds objects,
e.g. documents and terms or authors~\cite{doc,sci,movie}, or patient and gene~\cite{gene}.
Furthermore, real-world networks are often bipartite, and most unipartite networks
are projections of bipartite networks or exhibit bipartite properties~\cite{guillaume0,guillaume}.
In order to assist the visualization and navigation of large networks, one possibility is the use of
multilevel strategies, which consist on the employment of incremental coarsening of the original
network to obtain a sequence of simplified representations.
Multilevel strategies are most traditionally used for executing complex algorithms
on large-scale networks by the application of the algorithm on a smaller
version of the network~\cite{alan2,ml2}.
The employment of multilevel strategies for the visualization of simple (i.e. ``unipartite'')
networks have been reported, but their exploitation for visualizing bipartite networks
was not found by the authors, as described in the next subsection.
Accordingly, we present a system for the visualization of bipartite networks using
multilevel strategies developed for bipartite networks.
The system consists on presenting a simplified version of the network to the user, which then
requests for supervertices (or collections of them) to be uncoarsened and visualized in more detail.

This paper is organized as follows: in Section~\ref{rel} the related work is examined,
while in~\ref{nom} are selected remarks about the vocabulary.
The method is delineated in Section~\ref{des} and
the software implementation is then described in Section~\ref{sof}.
Results and discussion are in~\ref{res}.
Finally, Section~\ref{con} holds concluions and further work envisioned.

\subsection{Related work}\label{rel}
%%%
% artigos que devemos citar pelo histórico científico
% da estratégia ML, da visualização ML
% e os que devemos citar pelo ML em redes bipartidas
% e desenvolvidos no ICMC
Multilevel strategies have been employed to visualize unipartite networks~\cite{u1,u2,u3,u4,u5,u6,u7}.
Also, the aggregation of clusters have been reported, and comprises an approach that resembles
the coarsening procedure in creating simplified representations of the original network~\cite{a1,a2,a3,a4,a5,a6}.
Even so, the authors are not aware of previous reports on the use of multilevel strategies
for the visualization and navigation of bipartite networks.
Furthermore, only~\cite{a1,a2,a3,a4,a6} reports on a system suitable for visual analytics, i.e. the other works do not
address data analysis by interactive visual interfaces, including all multilevel strategies for visualization.

\subsection{Nomenclature remarks}\label{nom}
%%%
% level vs layer
% simple and heterogeneous graphs
% multiplex, ?
% multilevel strategies for optimization
% supernode metanode, 
The main vocabulary issue that arises in the context of this article is between
level and layer. Layers are the (two) node partitions of a bipartite network, while levels comprise the sequence
of coarsened networks.
Also, simple or homogeneous networks are opposed to heterogeneous networks, in which there
are more then one type of node.
Bipartite networks may be regarded as an elementary case of heterogeneous networks,
but it has one additional restriction: only nodes of different types are connected.
Another usual concept in this context is that of a multilayer network, in which there are layers,
i.e. partitions of nodes, in addition to nodes and links.
Bipartite networks, in this case, are multilayer networks with only two layers and no intra-layer links, i.e. all the links are inter-layers.
Most importantly, there are many synonyms which are found in this context, e.g. supernodes are also called metanodes or supervertices.
A thorough exposition of the vocabulary is beyond the scope of this article, but
some attention for the issue is helpful to assist searches and newcomers.

@@

A multilevel strategy consists in obtaining a hierarchy of coarsened networks $G_l$, $l$ integer and $l \in [0,L-1]$ where $G_0$ is the original network
and where $|V_i| \leq |V_{i+1}|$.
The coarsening procedures requires two algorithms, the \emph{matching}, that defines the nodes to be collapsed, and the \emph{contraction}, which builds the reduced representation
from the matched nodes.
There are several coarsening algorithms reported in the scientific literature~\cite{}, a few of them
developed for bipartite networks~\cite{}.
A discussion of these algorithms is found on the bibliography~\cite{} and is beyond the scope of this article.
Most importantly, we are interested in the MLPB algorithm, not only because it is suited for bipartite networks,
but because it has yielded simplified networks which carries essential topological features of the original networks,
and that result is perceived in the visualization as shown in Sections~\ref{des},~\ref{sof} and~\ref{res}.
